The Categorical Structure of Profunctor Optics

### 2017-08-03 paper.tex
.
Some thoughts on a weaker definition of optic

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{axiom}[theorem]{Axiom}
\newtheorem*{theoremstar}{Theorem}
\newtheorem{fact}[theorem]{Fact}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{convention}[theorem]{Convention}
\newtheorem{construction}[theorem]{Construction}
\newtheorem{example}[theorem]{Example}
\newtheorem{examples}[theorem]{Examples}
\newtheorem{notation}[theorem]{Notation}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{idea}[theorem]{Idea}
\newtheorem{question}[theorem]{Question}

@@

\todo{People to credit:
\begin{itemize}
\item The Haskell lens people: edwardk, shachaf, xplat, roconnor... (TODO: find out who these people are)
\item The Doubles paper
\item Bartosz's blog post which was written about the same time I started this. (Only tensor-like optics, no discussion of laws, confusion about enrichment(?))
\item The profunctor optics paper out of Oxford. (No generality, nothing categorical, no discussion of laws)
\end{itemize}
}

@@

\subsection{Notation}
\todo{I am open to changing any/all of these}
\begin{itemize}
\item Given $f : A \to B$ and $g : A \to C$, I write $[f, g] : A \to B \times C$ and $!_A : A \to 1$.
\item Given $f : A \to B$ and $g : C \to D$, I write $f \times g : A \times C \to B \times D$.
\item Given a $x \in P(C, C)$, I write $\langle x \rangle \in \int^C P(C, C)$ for its image, and $\copr_C : P(C,C) \to \int^C P(C, C)$
\end{itemize}
\section{Optics}

@@

Just as lenses are given by two concrete maps $\fget : S \to A$ and $\fput : S \times A \to S$, prisms are constructed by two maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\PreOptic_\sqcup((S, S), (A, A)) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a pre-prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = [r\inl, \id_S]l$

The laws for prisms are:
\begin{align*}
\fmatching \; \freview = \inr \\
[\id_S, \freview] \fmatching = \id_S
\end{align*}

\begin{proposition}
\label{prop-OpticImpliesLensLaws}
If $\langle l, r \rangle : S \hto A$ is a prism then the associated $\fmatching$ and $\freview$ functions satisfy the above laws.
\end{proposition}
\begin{proof}
\todo{Should be straightforward}
\end{proof}

@@

@@

\todo{The following original attempt is wrong, does not capture the right thing}

\begin{proposition}
Suppose the pre-lens $\langle l, r \rangle : S \hto A$ satisfies $rl = \id_S$ and $\langle lr \rangle = \langle t \times A \rangle$ as an element of $\int^{M \in \M} \C(M \times A, M \times A)$. Then $p$ obeys the $\fput\fget$ and $\fget\fput$ laws.
\end{proposition}
\begin{proof}
This is nearly identical to Proposition \ref{prop-OpticImpliesLensLaws}. As was mentioned in the proof of Proposition \ref{prop-onthenose}, we may assume that $lr = t \times A$ on the nose.

Now, again letting $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$,
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (t \times A) (\pi_1 l \times A) \\
&= \pi_2 (t \pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S
\end{align*}
\end{proof}

This suggests the following definition of merely well-behaved optics, which we may as well call semi-optics.

\begin{definition}
A \emph{semioptic} is a preoptic $p : S \hto A$ such that that $outside(p) = \id_S$ and $inside(p) = \langle t_A \rangle$ for some $t : M \to M$.
\end{definition}

Note that $t_A$ is necessarily idempotent, so if the ``evaluation-at-$A$'' functor is faithful then $t$ itself is idempotent.

\begin{proposition}
There is an subcategory $\SemiOptic_\M$ of $\PreOptic_\M$ given by objects of the form $(X, X)$ and semioptics between them.
\end{proposition}

@@

>>>>

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
%\newtheorem{axiom}[theorem]{Axiom}
\newtheorem*{theoremstar}{Theorem}
%\newtheorem{fact}[theorem]{Fact}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
%\newtheorem{convention}[theorem]{Convention}
%\newtheorem{construction}[theorem]{Construction}
\newtheorem{example}[theorem]{Example}
%\newtheorem{examples}[theorem]{Examples}
%\newtheorem{notation}[theorem]{Notation}
\newtheorem{remark}[theorem]{Remark}
%\newtheorem{idea}[theorem]{Idea}
%\newtheorem{question}[theorem]{Question}

@@

\todo{Past work:
\begin{itemize}
\item The Haskell lens people: edwardk, shachaf, xplat, roconnor... (TODO: find out who these people are)
\item The Doubles paper
\item Bartosz's blog post which was written about the same time I started this. (Only tensor-like optics, no discussion of laws, confusion about enrichment(?))
\item The profunctor optics paper out of Oxford. (No generality, nothing categorical, no discussion of laws)
\end{itemize}
}

@@

\subsection{Notation}
\todo{I am open to changing any/all of these}
\begin{itemize}
\item Given $f : A \to B$ and $g : A \to C$, I write $[f, g] : A \to B \times C$ and $!_A : A \to 1$.
\item Given $f : A \to B$ and $g : C \to D$, I write $f \times g : A \times C \to B \times D$.
\item Given a $x \in P(C, C)$, I write $\langle x \rangle \in \int^C P(C, C)$ for its image, and $\copr_C : P(C,C) \to \int^C P(C, C)$
\end{itemize}

\todo{
\subsection{(Co)ends and (Co)Yoneda}
}

\section{Optics}

@@

Just as lenses are constructed by two concrete maps $\fget : S \to A$ and $\fput : S \times A \to S$, prisms are constructed by two maps $\freview : A \to S$ and $\fmatching : S \to S \sqcup A$. These names are taken from the Haskell \texttt{lens} library.
\begin{align*}
\PreOptic_\sqcup((S, S), (A, A)) &= \int^{M \in \M} \C(S, M \sqcup A) \times \C(M \sqcup A, S) \\
&\cong \int^{M \in \M} \C(S, M \sqcup A) \times \C(M, S) \times \C(A, S) && \text{(universal property of coproduct)} \\
&\cong \C(S, S \sqcup A) \times \C(A, S) && \text{(co-Yoneda)}
\end{align*}
So, if we are given a pre-prism $\langle l, r \rangle : (S, S) \hto (A, A)$ then associated $\freview$ and $\fmatching$ morphisms are given by $\freview = r \inr$ and $\fmatching = (r\inl \sqcup \id_A)l$

The laws for prisms are:
\begin{align*}
\fmatching \; \freview = \inr \\
[\id_S, \freview] \fmatching = \id_S
\end{align*}
\todo{Are we missing a law?? Analogous to PutPut? I believe we are}
\todo{
\begin{proposition}
When $\C = \Set$, the third law is implied by the other two.
\end{proposition}
}

\begin{proposition}
\label{prop-OpticImpliesPrismLaws}
If $\langle l, r \rangle : S \hto A$ is a prism then the associated $\fmatching$ and $\freview$ functions satisfy the above laws.
\end{proposition}
\begin{proof}
\todo{Again it seems it is good enough if we only have $lr = t \sqcup A$}
\begin{align*}
\fmatching \; \freview 
&= (r\inl \sqcup \id_A)l r \inr \\
&= (r\inl \sqcup \id_A) \inr \\
&= \inr \\
[\id_S, \freview] \fmatching
&= [\id_S, r \inr] (r\inl \sqcup \id_A)l \\ 
&= [r\inl, r \inr] l \\ 
&= rl \\
&= \id_S
\end{align*}
\end{proof}

@@

\subsection{Change of Action}

\todo{
\section{Weaker Condition??}
}
\begin{proposition}
Suppose the pre-lens $\langle l, r \rangle : S \hto A$ satisfies $rl = \id_S$ and $\langle lr \rangle = \langle t \times A \rangle$ as an element of $\int^{M \in \M} \C(M \times A, M \times A)$. Then $p$ obeys the three lens laws.
\end{proposition}
\begin{proof}
This is nearly identical to Proposition \ref{prop-OpticImpliesLensLaws}. As was mentioned in the proof of Proposition \ref{prop-onthenose}, we may assume that $lr = t \times A$ on the nose.

Now, again letting $\fget = \pi_2 l$ and $\fput = r (\pi_1 l \times A)$,
\begin{align*}
\fget \; \fput 
&= \pi_2 l r (\pi_1 l \times A) \\
&= \pi_2 (t \times A) (\pi_1 l \times A) \\
&= \pi_2 (t \pi_1 l \times A) \\
&= \pi_2 \\
\fput [\id_S, \fget] 
&= r (\pi_1 l \times A) [\id_S, \pi_2 l] \\
&= r [\pi_1 l, \pi_2 l] \\
&= rl \\
&= \id_S \\
\fput (\fput \times A) 
&= r (\pi_1 l \times A) (r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 l  r (\pi_1 l \times A) \times A) \\
&= r (\pi_1 (t\pi_1 l \times A) \times A) \\
&= r (t\pi_1 l \times A) \pi_{1,3} \\
&= r lr(\pi_1 l \times A) \pi_{1,3} \\
&= r(\pi_1 l \times A) \pi_{1,3} \\
&= \fput \, \pi_{1, 3}
\end{align*}
\end{proof}

\begin{proposition}
Suppose there exists a map $x : 1 \to A$. Then if $\fget : S \to A$ and $\fput : S \times A \to S$ satisfy the three lens laws then the corresponding $p : S \hto A$ obeys the above weaker condition.
\end{proposition}
\begin{proof}
Recall that the corresponding optic $p : S \hto A$ has representative $\langle [\id_s, \fget], \fput \rangle$.

Let $\fput_x : S \to S$ be the composite $S \to S \times 1 \xrightarrow{x} S \times A \xrightarrow{\fput} S$. The commutative diagram
\[
\begin{tikzcd}
S \times A \ar[r, "\fput_x \times A "] \ar[d, "lr", swap] & S \times A \ar[d, "\fput_x \times A"] \ar[dl, "lr"] \\
S \times A \ar[r, "\fput_x \times A", swap] & S \times A
\end{tikzcd}
\]
proves that $\langle lr \rangle = \langle \fput_x \times A \rangle$ in $\int^{M \in \M} \C(M A, M A)$.
\todo{Is there some property of $\times$ used here we can abstract?}
\end{proof}

@@

@@

Some thoughts: obvious thing to try is a double category where objects are the objects of $\C \times \C$, vertical morphisms $(S,T) \to (A, B)$ are maps $S \to A$ and $B \to T$, horizontal morphisms are preoptics. This I think is framed, you can always turn a pair of maps into a preoptic.

Note the pair of maps above is also the set of morphisms $E_{S, T} \to E_{A, B}$. Maybe the objects should be profunctors??
